Bees need larger brains to thrive in urban environments

### 2023-09-06 Manuscript/Ms.Rmd
Minor update on ms (fig 1 caption)
Response to R

---
title: "Bees need larger brains to thrive in urban environments"
output:
  pdf_document:
    fig_caption: yes
  word_document: default
link-citations: yes
linkcolor: RoyalBlue
urlcolor: RoyalBlue
fontsize: 12pt
bibliography: references.bib
always_allow_html: true
header-includes:
- \usepackage{orcidlink}
- \usepackage{caption,setspace}
- \captionsetup[figure]{labelformat=empty}
- \usepackage{float}
- \floatplacement{figure}{H}
csl: BiologyLetters.csl
---

@@

The rapid conversion of natural habitats to anthropogenic landscapes is threatening insect pollinators worldwide, raising concern on the consequences for their fundamental role as plant pollinators. However, not all pollinators are negatively affected by habitat conversion, as certain species find in anthropogenic landscapes appropriate resources to persist and proliferate. The reason why some species thrive in anthropogenic environments while most find them inhospitable remains poorly understood. The cognitive buffer hypothesis, widely supported in vertebrates but untested in insects, offers a potential explanation. This theory suggests that species with larger brains have enhanced behavioural plasticity, enabling them to confront and adapt to novel challenges. To investigate this hypothesis in insects, we measured brains for `r nrow(raw)` individuals from `r nlevels(factor(unique(raw$Species)))` bee species, and evaluated the association of their brain sizes with detailed information on habitat preferences. Our analyses revealed that bees that prefer urban habitats had larger brains relative to their body size than those who prefer forested or agricultural habitats. Additionally, urban bees exhibited larger body sizes and, consequently, larger absolute brain sizes. Our results provide the first empirical support for the cognitive buffer hypothesis in invertebrates, suggesting that a large brain in bees could confer behavioural advantages to tolerate urban environments.

@@

Pollinators deliver a fundamental ecosystem service on which the earth's vegetation and human economy depend [@ollerton2017]. Regrettably, there is increasing evidence of recent declines in pollinator populations [@bartomeus2019; @potts2010; @scheper2014]. One of the main contributing factors to the current pollinator declines is the alteration and loss of their habitat due to human activity [@harrison2015; @harrison2018]. Anthropogenic landscapes present new challenges for the survival and reproduction of organisms, increasing their risk of extinction by maldaptation [@winfree2011; @brown2009 ; @harrison2019]. Yet, not all pollinator species are negatively affected by land use change. For instance, some bee species are able to tolerate human-altered environments [@biesmeijer2006; @bartomeus2013; @collado2019] or even to thrive [e.g., *Bombus terrestris*\; @rasmont2008; @theodorou2020].

Human-dominated habitats ---notably cities--- drastically modify the original conditions where pollinators evolved, but can also offer unique ecological opportunities in the form of new nesting spots, reduced predation pressure and high food availability associated with non-indigenous plants [@cane2006; @winfree2011; @collado2019]. The question is then why only some species are able to tolerate urban environments. The "cognitive buffer hypothesis" provides an explanation for this conundrum, suggesting that in novel environments the chances to successfully survive and reproduce depends on cognition [@sol2009], that is, the processes involved in gathering, storing and reacting to environmental information [@shettleworth2009]. While the "cognitive buffer hypothesis" receives ample support from studies of vertebrates [@sol2008; @amiel2011; @sayol2016; @howell2023], similar evidence is lacking for insects.

Insects have been long associated with the most basic types of learning due to their miniature brain [@giurfa2015; @perry2017]. However, numerous recent studies have shown that bees and other insects have sophisticated cognition that goes beyond simple associative learning or conditioning [@chittka2009; @avargues2011; @sheehan2011]. Some of these more complex forms of cognition involve the use of simple tools, attention, social learning or metacognitive processes [@giurfa2015; @perry2017]. In addition, there is also evidence for substantial variation across species in brain size---both in absolute terms and relative to body size--- with species that have larger brains also exhibiting enhanced cognitive performance [@collado2021]. Therefore, we can ask whether the varying success of insects in human-altered habitats might be explained by variation in brain size.

@@

Our dataset contains measurements of brain and body size for bee specimens captured on flowers by hand netting in different areas of the East Coast of the United States and Europe (Spain and the Netherlands). The dataset includes information of `r nrow(raw)` female individuals from `r nlevels(factor(unique(raw$Species)))` species that represent `r nlevels(factor(unique(tax$family)))` families and `r nlevels(factor(unique(tax$genus)))` genera. We only considered female specimens because: (i) they are involved in a greater diversity of tasks than males, and hence are expected to experience greater environmental pressures; and (ii), have brains that are structurally and functionally different from those of males [@streinzer2013; @roselino2015]. Brain size was measured as the weight of fixed brains [@sayol2020] and body size as intertegular span [@kendall2019; see supplementary text]. Given that brain size scales allometrically with body size [@sayol2020], we considered for our analysis the size of the brain relative to the body.

@@

## (b) Habitat preferences

@@

We downloaded occurrence information for all the measured species from the Global Biodiversity Information Facility (GBIF; <http://www.gbif.org/>) for North America and Europe. The data was downloaded through R programming language with the help of the function *occ_download* from the package *rgbif* version 3.7.3 [@chamberlain2017]. We selected the states or countries with the highest density of records for our set of species. For North America, we selected states located on the East Coast of the United States (Figure S2A), covering an approximate area of `r usakm2` km$^{2}$. For Europe, we selected countries located on the north and centre of the continent (Figure S2B), representing a total area of `r eukm2` km$^{2}$. To further reduce biases in the data, we only included species with a minimum number of 50 records and whose geographic distribution was larger than the sampled area (i.e., excluding species at the edge of their distributions). In addition, we optimised the match between species occurrence and the land cover data by only using georeferenced records obtained between 1990 and 2022 with a minimum of two decimals of latitude/longitude coordinates.

We assigned a habitat type to each GBIF occurrence by merging land cover information with the georeferenced records of species occurrences. The land cover classification was obtained from the 2006 online inventories of the National Land Cover Database (NLCD) for United States and the Corine Land Cover (CLC) for Europe. After downloading these inventories as raster files, we used the functions *rast* and *extract* from the *Terra* package version 1.6-41 [@hijmans2022] to read and obtain the cover classification of the different georeferenced records, respectively. To simplify the interpretation and conduct a joint analysis for both regions, we divided the resulting cover classes into three single categories: (i) natural, (ii) agricultural and (iii) urban (see Tables S1 and S2 for details). With this information we can build an occurrence matrix with species in rows, habitats in columns, and cells depicting the number of occurrences per species-habitat combination.

@@

To evaluate how the association between habitat preference and relative brain size changed by habitat type, we modelled with Bayesian generalised linear models their association. For this, we first joined the resulting habitat preference and relative brain size datasets by species. The resulting distribution of habitat preferences for each of the habitats analysed followed a zero-one inflated beta distribution (Figure S3), indicating that there were high frequencies of habitat preferences close to 0 or 1 but low frequencies of intermediate values between 0 and 1. Hence, in our analyses we take a conservative approach and only modelled the extremes of the distribution (i.e., species classified as avoiding or preferring a given habitat). Because we assessed habitat preference as binary ("avoiding" or "preferring"), we specified a Bernouilli distribution where habitat preference was the response variable and relative brain size the predictor. Again, we included the phylogenetic covariance matrix as a random factor. Moreover, we also explored how habitat preference changed by brain weight and intertegular span independently. For this, we conducted two analogous models with these two different predictors. In addition, we also investigated the different trends for the United States and Europe separately.

@@

Habitat preference varied substantially across species (Figure 1B) and showed moderate phylogenetic signal (i.e., preference for natural habitat; $\lambda$ = `r lambda_nat`; *P* = `r p_nat`). In general, most bee species preferred one or two habitat types but more rarely occurred indistinctly in the three habitats (Figure 1B). The most preferred habitats were the agricultural and natural ones with `r pref_agr` and `r pref_nat` species over 80% of the values from null models, respectively (Figure 1B and Figure S3). In contrast, most species avoided urban habitats (`r pref_urb1` species under the 20th percentile) and just `r pref_urb2` species showed high preference for this habitat type (Figure 1B and Figure S3).

```{r, echo=FALSE, message=FALSE, cache=FALSE, warning=FALSE,fig.align="center", fig.height=12, fig.width=12,fig.cap="Figure 1. Phylogenetic relationship and habitat preferences for the selected bee species with brain weight and intertegular distance information (N=89). A) Phylogenetic tree at the genus level. Tree branches are coloured based on the geographical location of the different bee species (North of Europe, East Coast of the United States or from both regions). The deviation of the brain size in relation to the body (i.e., residuals) is indicated with weighted points at the end of the tip branches. Larger points indicate larger brains in proportion to their body size and vice versa. B) Heatmap showing the habitat preference for each bee species. The columns delimit the habitat type (i.e., natural, agricultural and urban) and the rows the different bee species."}

@@

We found that bee species relative brain size is associated with the habitat type preference (Figure 2A; Bayesian R$^2$ = `r r2_main`). Specifically, we found that bees with larger brains relative to body size showed a higher preference for urban habitats than bees with smaller relative brain sizes (Figure 2A). Contrarily, bees with smaller relative brains showed higher preference in natural and agricultural habitats than bees with larger relative brain sizes (Figure 2A). The models of the association between absolute brain size and intertegular span with habitat preference also showed a marked differences between habitat types (Figures 2B and 2C; Bayesian R$^2$ = `r r2_brain`; Bayesian R$^2$ = `r r2_its`, respectively). Specifically, we found that bees with larger brains and body sizes tend to appear more in urban habitats and bees with smaller brains and body sizes are more common in natural and agricultural ones (Figures 2B and 2C). These findings were consistent with the analogous analyses separated by geographical regions (United States and Europe; Figure S4).

@@

Using data from two different continents and `r nlevels(factor(unique(raw$Species)))` bee species, we find that bees with a preference for inhabiting urban habitats tended to have larger brains relative to their body while those with forest or agricultural preferences had relatively smaller brains. These results are in line with the cognitive buffer hypothesis [@sol2009], which predicts that a large relative brain should provide enhanced behavioural plasticity to persist and thrive in novel environments. In urban environments, individuals are frequently exposed to a variety of new challenges such as novel resources or human disturbances that can change fast in time and space [@wheater2002]. In such scenarios, a large brain could confer the cognitive flexibility needed to efficiently exploit novel resources while avoiding risks. While the exact mechanisms are unclear, we expect cognitive flexibility to be essential in a variety of contexts such as the use of human-made materials for the nest [@allasino2019] or the need to collect pollen and nectar from exotic flowers [@lowenstein2019].

Also in line with the cognitive buffer hypothesis [@ducatez2020], we find that bees that proliferate in urban habitats have bigger bodies and larger absolute brains. Larger species require longer development times and obtain greater net benefits from exploring and learning, especially in heterogeneous environments [@bateman2012; @theodorou2021]. For example, the large carpenter bees of the genus _Xylocopa_ can live up to two years and are frequent urban dwellers, while small _Andrena_ forest specialists complete their adult life cycle in a few weeks. The patchiness of urban resources also seems to favour larger body sizes, as suggests the observation that foraging distance is positively associated with body size in bees [@greenleaf2007; @kendall2019]. There are also shreds of evidence that associate increased tolerance to urbanisation with wider ecological niches [@geslin2013; @buchholz2020; @prendergast2022], and both larger bodies and relative brains are thought to be features that can facilitate broader diets and niche expansion. Further understanding all these processes represent an important avenue of future research.

Overall, our findings support and expand previous evidence found in vertebrates that having a large brain size can allow survival in novel environments [@santini2019; @sayol2020b], highlighting that a cognitive buffer is possible even with tiny brains. The use of the entire brain size as a proxy for cognitive performance is not exempt from criticism [@chittka2009; @lihoreau2012]. Our analyses focused on brain size mainly for reasons of data availability. The use of brain size is justified by the existence of previous evidence that enlarged brains enhance cognitive performance in bees [@collado2021]. Moreover, brain size is less subject to measurement error or context-dependent biases in comparison with other experimental measures of cognition [@chittka2009; @lihoreau2012]. Although the use in the future of finer measures like neuropils size or mushroom bodies [the suspected centres of cognitive processes in bees\; @barth1997; @withers1995; @fahrbach1995] is likely to help improve our understanding of bee cognitive abilities. Our findings highlight the importance of behavioural responses for understanding the dynamics of insect populations in altered environments and stresses the need to avoid viewing them as passive agents of external pressures.

@@

We thank Curro Molina, Ivo Reamakers and Parker Gambino for collecting some of the studies specimens.

>>>>

---
title: "Bees need larger brains to thrive in urban environments"
output:
  pdf_document:
    fig_caption: yes
  word_document: default
link-citations: yes
linkcolor: RoyalBlue
urlcolor: RoyalBlue
fontsize: 12pt
includeline-num: true
bibliography: references.bib
always_allow_html: true
header-includes:
- \usepackage{orcidlink}
- \usepackage{caption,setspace}
- \captionsetup[figure]{labelformat=empty}
- \usepackage{float}
- \floatplacement{figure}{H}
- \usepackage[left]{lineno}
- \linenumbers
csl: BiologyLetters.csl
---

@@

The rapid conversion of natural habitats to anthropogenic landscapes is threatening insect pollinators worldwide, raising concern on the consequences for their fundamental role as plant pollinators. However, not all pollinators are negatively affected by habitat conversion, as certain species find in anthropogenic landscapes appropriate resources to persist and proliferate. The reason why some species thrive in anthropogenic environments while most find them inhospitable remains poorly understood. The cognitive buffer hypothesis, widely supported in vertebrates but untested in insects, offers a potential explanation. This theory suggests that species with larger brains have enhanced behavioural plasticity, enabling them to confront and adapt to novel challenges. To investigate this hypothesis in insects, we measured brains for `r nrow(raw)` individuals from `r nlevels(factor(unique(raw$Species)))` bee species, and evaluated the species-level association between brain size and habitat preferences. Our analyses revealed that bee species that prefer urban habitats had larger brains relative to their body size than those who prefer forested or agricultural habitats. Additionally, urban bees exhibited larger body sizes and, consequently, larger absolute brain sizes. Our results provide the first empirical support for the cognitive buffer hypothesis in invertebrates, suggesting that a large brain in bees could confer behavioural advantages to tolerate urban environments.

@@

Pollinators deliver a fundamental ecosystem service on which the earth's vegetation and human economy depend [@ollerton2017]. Regrettably, there is increasing evidence of recent declines in pollinator populations [@bartomeus2019; @potts2010; @scheper2014]. One of the main contributing factors to the current pollinator declines is the alteration and loss of their habitat due to human activity [@harrison2015; @harrison2018]. Anthropogenic landscapes present new challenges for the survival and reproduction of organisms, increasing their risk of extinction by maladaptation [@winfree2011; @brown2009 ; @harrison2019]. Yet, not all pollinator species are negatively affected by land use change. For instance, some bee species are able to tolerate human-altered environments [@biesmeijer2006; @bartomeus2013; @collado2019] or even to thrive [e.g., *Bombus terrestris*\; @rasmont2008; @theodorou2020].

Human-dominated habitats ---notably cities--- drastically modify the original conditions where pollinators evolved, but can also offer unique ecological opportunities in the form of new nesting spots, shelter from phytosanitary products, reduced predation pressure and high food availability associated with non-indigenous plants [@cane2006; @winfree2011; @collado2019]. The question is then why only some species are able to tolerate and exploit urban environments. The "cognitive buffer hypothesis" provides an explanation for this conundrum, suggesting that in novel environments the chances to successfully survive and reproduce depends on cognition [@sol2009], that is, the processes involved in gathering, storing and reacting to environmental information [@shettleworth2009]. While the "cognitive buffer hypothesis" receives ample support from studies of vertebrates [@sol2008; @amiel2011; @sayol2016; @howell2023], similar evidence is lacking for insects.

Insects have been long associated with the most basic types of learning due to their miniature brain [@giurfa2015; @perry2017]. However, numerous recent studies have shown that bees and other insects have sophisticated cognition that goes beyond simple associative learning or conditioning [@chittka2009; @avargues2011; @sheehan2011]. Some of these more complex forms of cognition involve the use of simple tools, attention, social learning or metacognitive processes [@giurfa2015; @perry2017]. In addition, there is also evidence for substantial variation across species in brain size---both in absolute terms and relative to body size--- with species that have larger brains also exhibiting enhanced cognitive performance, at least for some tasks such as conditional learning [@collado2021]. Therefore, we can ask whether the varying success of insects in human-altered habitats might be explained by variation in brain size.

@@

Our dataset contains measurements of brain and body size for bee specimens captured on flowers by hand netting in different areas of the East Coast of the United States and Europe (Spain and the Netherlands). These specimens were collected opportunistically mostly in semi-natural habitats. The dataset includes information of `r nrow(raw)` female individuals from `r nlevels(factor(unique(raw$Species)))` species that represent `r nlevels(factor(unique(tax$family)))` families and `r nlevels(factor(unique(tax$genus)))` genera. We only considered female specimens because: (i) they are involved in a greater diversity of tasks than males, and hence are expected to experience greater environmental pressures; and (ii), have brains that are structurally and functionally different from those of males [@streinzer2013; @roselino2015]. Brain size was measured as the weight of fixed brains [@sayol2020] and body size as intertegular span [@kendall2019; see supplementary text S1]. Given that brain size scales allometrically with body size [@sayol2020], we considered for our analysis the size of the brain relative to the body.

@@

## (b) Diet specialisation

To ensure diet is not acting as a confounding factor we investigated how diet specialisation is associated with both habitat preference and brain size. For this, we used the diet information collected in @sayol2020 and complemented the missing information by using the same online resources used in @sayol2020. Bee species were classified as oligolectic when they use a single plant family to feed their brood or polylectic when they use several.  We tested for statistical differences in diet specialization per habitat type preference and brain size (i.e., both absolute and relative brain size) by using the Wilcoxon test. We found a low number of specialist species in our dataset with no clear associations with habitat preference or brain size that can explain the relationship between habitat preference and brain size (see details in Figure S2).

## (c) Habitat preferences

@@

We downloaded occurrence information for all the measured species from the Global Biodiversity Information Facility (GBIF; <http://www.gbif.org/>) for North America and Europe. The data was downloaded through R programming language with the help of the function *occ_download* from the package *rgbif* version 3.7.3 [@chamberlain2017]. We selected the states or countries with the highest density of records for our set of species. For North America, we selected states located on the East Coast of the United States (Figure S3A), covering an approximate area of `r usakm2` km$^{2}$. For Europe, we selected countries located on the north and centre of the continent (Figure S3B), representing a total area of `r eukm2` km$^{2}$. To further reduce biases in the data, we only included species with a minimum number of 50 records and whose geographic distribution was larger than the sampled area (i.e., excluding species at the edge of their distributions). In addition, we optimised the match between species occurrence and the land cover data by only using georeferenced records obtained between 1990 and 2022 with a minimum of two decimals of latitude/longitude coordinates.

We assigned a habitat type to each GBIF occurrence by merging land cover information with the georeferenced records of species occurrences. The land cover classification was obtained from the 2006 online inventories of the National Land Cover Database (NLCD) for the United States and the Corine Land Cover (CLC) for Europe. After downloading these inventories as raster files, we used the functions *rast* and *extract* from the *Terra* package version 1.6-41 [@hijmans2022] to read and obtain the cover classification of the different georeferenced records, respectively. To simplify the interpretation and conduct a joint analysis for both regions, we divided the resulting cover classes into three single categories: (i) natural, (ii) agricultural and (iii) urban (see Tables S1 and S2 for details). With this information we can build an occurrence matrix with species in rows, habitats in columns, and cells depicting the number of occurrences per species-habitat combination.

@@

Our macro-ecological approach is justified by: (i) independent data that ensures that any ecological patterns, if present, are likely to be robust and generalizable; (ii) low intraspecific variation of brain and body size in comparison to the interspecific variation (See supplementary text S2 and Figure S4) that provides biological meaning of averaging relative brain sizes per species and the study of the ecological patterns across them.

To evaluate how the association between habitat preference and relative brain size per species changed by habitat type, we used a Bayessian approach to model their association. For this, we first joined the resulting habitat preferences per species with their respective average relative brain sizes. The resulting distribution of habitat preferences for each of the habitats analysed followed a zero-one inflated beta distribution (Figure S5), indicating that there were high frequencies of habitat preferences close to 0 or 1 but low frequencies of intermediate values between 0 and 1. Hence, in our analyses we take a conservative approach and only modelled the extremes of the distribution (i.e., species classified as avoiding or preferring a given habitat). Because we assessed habitat preference as binary ("avoiding" or "preferring"), we specified a Bernouilli distribution where habitat preference was the response variable and relative brain size the predictor. Again, we included the phylogenetic covariance matrix as a random factor. Moreover, we also explored how habitat preference changed by average brain weight and intertegular span independently. For this, we conducted two analogous models with these two different predictors. In addition, we also investigated the different trends for the United States and Europe separately.

@@

Habitat preference varied substantially across species (Figure 1B) and showed moderate phylogenetic signal (i.e., preference for natural habitat; $\lambda$ = `r lambda_nat`; *P* = `r p_nat`). In general, most bee species preferred one or two habitat types but more rarely occurred indistinctly in the three habitats (Figure 1B). The most preferred habitats were the agricultural and natural ones with `r pref_agr` and `r pref_nat` species over 80% of the values from null models, respectively (Figure 1B and Figure S5). In contrast, most species avoided urban habitats (`r pref_urb1` species under the 20th percentile) and just `r pref_urb2` species showed high preference for this habitat type (Figure 1B and Figure S5).

```{r, echo=FALSE, message=FALSE, cache=FALSE, warning=FALSE,fig.align="center", fig.height=12, fig.width=12,fig.cap="Figure 1. Phylogenetic relationship and habitat preferences for the selected bee species with brain weight and intertegular distance information (N=89). A) Phylogenetic tree at the genus level. Tree branches are coloured based on the geographical location of the different bee species (North of Europe, East Coast of the United States or from both regions). The deviation of the brain size in relation to the body (i.e., residuals) is represented with filled circles of proportional area to the residuals at the end of the tip branches. Larger circles indicate larger brains in proportion to their body size and vice versa. B) Heatmap showing the habitat preference for each bee species. The columns delimit the habitat type (i.e., natural, agricultural and urban) and the rows the different bee species."}

@@

We found that bee species relative brain size is associated with the habitat type preference (Figure 2A; Bayesian R$^2$ = `r r2_main`). Specifically, we found that bees with larger brains relative to body size showed a higher preference for urban habitats than bees with smaller relative brain sizes (Figure 2A). Contrarily, bee species with smaller relative brains showed higher preference in natural and agricultural habitats than bees with larger relative brain sizes (Figure 2A). The models of the association between absolute brain size and intertegular span with habitat preference also showed a marked differences between habitat types (Figures 2B and 2C; Bayesian R$^2$ = `r r2_brain`; Bayesian R$^2$ = `r r2_its`, respectively). Specifically, we found that bees with larger brains and body sizes tend to appear more in urban habitats and bees with smaller brains and body sizes are more common in natural and agricultural ones (Figures 2B and 2C). These findings were consistent with the analogous analyses separated by geographical regions (United States and Europe; Figure S6).

@@

Using data from two different continents and `r nlevels(factor(unique(raw$Species)))` bee species, we find that bee species with a preference for inhabiting urban habitats tended to have larger brains relative to their body while those with forest or agricultural preferences had relatively smaller brains. These results are in line with the cognitive buffer hypothesis [@sol2009], which predicts that a large relative brain should provide enhanced behavioural plasticity to persist and thrive in novel environments. In urban environments, individuals are frequently exposed to a variety of new challenges such as novel resources or human disturbances that can change fast in time and space [@wheater2002]. In such scenarios, a large brain could confer the cognitive flexibility needed to efficiently exploit novel resources while avoiding risks. While the exact mechanisms are unclear, we expect cognitive flexibility to be essential in a variety of contexts such as the use of human-made materials for the nest [@allasino2019] or the need to collect pollen and nectar from exotic flowers [@lowenstein2019].

Also in line with the cognitive buffer hypothesis [@ducatez2020], we find that bees that proliferate in urban habitats have bigger bodies and larger absolute brains. Larger species require longer development times and obtain greater net benefits from exploring and learning, especially in heterogeneous environments [@bateman2012; @theodorou2021]. For example, the large carpenter bees of the genus _Xylocopa_ can live up to two years and are frequent urban dwellers, while small _Andrena_ forest specialists complete their adult life cycle in a few weeks. The patchiness of urban resources also seems to favour larger body sizes, as suggests the observation that foraging distance is positively associated with body size in bees [@greenleaf2007; @kendall2019]. There are also shreds of evidence that associate increased tolerance to urbanisation with wider ecological niches [@geslin2013; @buchholz2020; @prendergast2022], and both larger bodies and relative brains are thought to be features that can facilitate broader diets and niche expansion. Only a few bee specialists are analyzed in this paper, and further understanding how different life styles facilitate urban exploitation represent an important avenue of future research.

Overall, our findings support and expand previous evidence found in vertebrates that having a large brain size can allow survival in novel environments [@santini2019; @sayol2020b], highlighting that a cognitive buffer is possible even with tiny brains. The use of the entire brain size as a proxy for cognitive performance is not exempt from criticism [@chittka2009; @lihoreau2012]. Our analyses focused on brain size mainly for reasons of data availability. The use of brain size is justified by the existence of previous evidence that enlarged brains enhance some aspects of cognitive performance in bees, such as learning [@collado2021]. Moreover, brain size is less subject to measurement error or context-dependent biases in comparison with other experimental measures of cognition [@chittka2009; @lihoreau2012]. Although the use in the future of finer measures like neuropils size or mushroom bodies [the suspected centres of cognitive processes in bees\; @barth1997; @withers1995; @fahrbach1995] is likely to help improve our understanding of bee cognitive abilities. Further, there are other alternative explanations beyond the cognitive buffer hypothesis for the enlargement of brains in urban environments. Here, we only investigated the role of diet specialisation as a potential confounding factor but other ecological and physiological factors cannot be discarded as potential drivers of this pattern. Downscaling our analysis to the individual specimen level where brain size, habitat use and cognitive performance can be tracked through their lifespan would be a challenging but promising next step. Overall, our findings highlight the importance of behavioural responses for understanding the dynamics of insect populations in altered environments and stresses the need to avoid viewing them as passive agents of external pressures.

@@

We thank Fransico P. Molina, Ivo Reamakers and Parker Gambino for collecting some of the studies specimens.
